Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-07-21 analysis/paper/paper.Rmd
Setting up GOF with 4-CPL

# Introduction
Variation in population patterns are regarded important drivers and consequences of cultural variation, and of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2020]. During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population size and density [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].

@@

# Background
Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. method has a long tradition in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites when temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.  Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jørgensen2020].

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to consitute a 'site'. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @grocutt2020; @crema2022 for a review].

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that visual inspections of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2020. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model. These null-models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of regression or maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for interpreting if local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].

@@

```{r shoredate, echo = FALSE, fig.cap = "A) An example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.  B) The geological displacement curves in the region and the curve interpolated to the example site. C) The black line on the y-axis represents the elevation of the site. This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea, giving a \\textit{terminus post quem} date. D) The shoreline dating approach used for this study, where the gamma distrbution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of 0.108, and could therefore not be rejected. Based on the available data, this model represents the best explanation of the data. For the SSPD all of the standard models could be rejected.

@@

It is established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD starts from c. 9500 BCE, the RSPD show no evidence of human activity until c. 8500–8300 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs for fuel or use of rapidly deteriorating seal blubber for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [e.g. @damm2022]. While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind the observed pattern.

The considered models indicate that after the initial jumps, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. An immediate explanation of this mismatch could be that the SSPD largely reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers. A common interpretation of the Early Mesolithic in the region is that the population was low in numbers but highly mobile, reflected in large numbers of short lived sites along the coast [e.g. @bjerck2017]. This is commonly seen in relation to a rapid colonisation of the Norwegian coastline and homogeneous lithic technology over large parts of Scandinavia in this period. High mobility has been seen as necessary both to maintain demographic viability among colonising groups and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. At the same time, the low/lacking signal in the RSPD for this earliest period could be taken as evidence for low population numbers during the first settlement period following the deglaciation.

@berg-hansen2022 have argued that there is an increase in sites between c. 8000--7000 BCE in the northern part of the study area of this paper. However, their analysis covers a limited spatial area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021]. In contrast to their findings, the analysis undertaken here instead indicates a rapid increase followed drop in the number of sites through this time-interval, possibly corresponding with a shift to reduced residential mobility [@roalkvam2022; @damlien2018b; cf. @binford1980].

As evidenced by both genetic data and technological analyses of lithic inventories, the initial phase of human colonisation from the south appears to be followed by an influx, and possible mixing, of people from the east at the transition to the Middle Mesolithic period around 8300 BCE [@gunther2018; @manninen2021]. During the first part of the Middle Mesolithic, it has been suggested that the mobility patterns shifted to a more stable settlement, confined to larger regions [@damlien2018b]. If this narrative is coupled with the data observed here, these migratory events would appear to result in a relatively sudden drop in site-count as indicated in the SSPD. The end of this drop corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes rather than taphonomic factors.

@@

While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022]. Environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g. @jorgensen2020; @ordonez2022]. Furthermore, mapping and constrasting these developments across a large region will offer important contextualisation for the developments observed here.

>>>>

# Introduction
Variation in population patterns are regarded important drivers and consequences of cultural variation, and of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2021]. During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population size and density [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].

@@

# Background
Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. method has a long tradition in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites when temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.  Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jorgensen2020].

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to consitute a 'site'. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2020; @crema2022 for a review].

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model. These null-models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].

@@

```{r shoredate, echo = FALSE, fig.cap = "A) An example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.  B) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site. C) The black line on the y-axis represents the elevation of the site. This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea, giving a \\textit{terminus post quem} date. D) The shoreline dating approach used for this study, where the gamma distrbution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of 0.108, and could therefore not be rejected. Based on the available data, this model therefore represents the best explanation of the data. Given the failure to reject the logistic model, the subsequent step of also including a comparison with CPL models was not strictly necessary, but the result of this procedure further substantiates the finding as the logistic model is the best among the considered candidates.

For the SSPD all of the standard models could be rejected. The result of subjecting the SSPD to the ADMUR workflow identified 4-CPL as the best model. An important caveat here is that models with a complexity beyond 4-CPL consistently failed to converge. A fuller reporting of this convergence issue, including the exploration of different parameter settings, different sample sizes and alternative optimization algorithms is given in the supplementary material. Subjecting the 4-CPL model to the goodness-of-fit test gave p =, meaning that the data nonetheless could be expected under this model. The accepted models that underlies the comparison to follow are therefore the logistic model for the RSDP and the 4-CPL for the SSPD.

@@

It is established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD starts from c. 9500 BCE, the RSPD show no evidence of human activity until c. 8500–8300 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs for fuel or use of rapidly deteriorating seal blubber for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [e.g. @damm2022]. While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump at the start of RSPD, combined with the magnitude of the SSPD in the period preceding this, suggest that cultural rather than environmental factors are central drivers behind the observed pattern.

The considered models indicate that after the initial jumps, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. An immediate explanation of this mismatch could be that the SSPD mainly reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers. A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short lived sites along the coast [e.g. @bjerck2017]. This is commonly seen in relation to homogeneous lithic technology over large parts of Scandinavia in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline. High mobility has been seen as necessary both to maintain demographic viability among colonising groups and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. At the same time, the lacking signal in the RSPD for this earliest period could be taken as evidence for low population numbers during the first settlement period following the deglaciation.

@berg-hansen2022 have argued that there is an increase in sites between c. 8000--7000 BCE in the northern part of the study area of this paper. However, their analysis covers a limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].

In contrast to the findings of @berg-hansen2022, the analysis undertaken here instead indicates a drop in the number of sites through this time-interval. It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [@roalkvam2022; @damlien2018b]. Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east at the transition to the Middle Mesolithic period around the same time [@gunther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events would appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes rather than taphonomic factors.

@@

While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022]. Environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g. @jorgensen2020; @ordonez2022]. Furthermore, mapping and constrasting these developments across a large region will offer important contextualisation for the developments observed here.
